\chapter{Weightings}

First off, I need to get the 90\% bread-and-butter done extremely
well. This includes what the programmer faces each time they sit down
and start typing, and it includes the key benefits that are expected
from Cottontail.

Next, I don't want to waste all of my complexity on silly, whimsical,
half-ass features. There are limits on the total code size, and
there's limits on each component.

Finally, some parts of the language must be taught extremely well, and
others can be explorered and learned outside of this formal tutorial
document. These priorities need to be set too.


\section{Polish}

\begin{itemize}
\item Syntax:             40\%
\item Documentation:      30\%
\item REPL:               20\%
\item Machine code:       10\%
\end{itemize}


\section{Code size}

Here are some example code sizes:

\begin{itemize}
\item Linux:          9,400,000
\item GCC:            2,600,000
\item GNU C Library:  1,000,000
\item CPython:          660,000
\item Racket:           660,000
\item LLVM:             540,000
\item GLib:             340,000
\item GHC:              330,000
\item Go:               310,000
\item Antlr:             42,000
\item LuaJIT:            40,000
\item BoehmGC:           29,000
\item Readline:          24,000
\item Lua:               20,000
\item Tinyscheme:         5,000
\end{itemize}

There's no way I can write or maintain a behemoth like Linux or GCC.
Even Racket and Haskell are an order of magnitude too large for one
person, and for my performance goals. My total code size will be less
than LuaJIT's: 40,000 lines of code (including tests) is the limit.

In specific terms, my budget will be used as follows:

\begin{itemize}
\item Standard library:  10,000
\item Tests:             10,000
\item Typing:             5,000
\item Compilation:        5,000
\item REPL:               3,000
\item Hot swapping:       3,000
\item Parsing:            2,000
\item Annotations:        2,000
\end{itemize}


\section{Formal documentation}

\begin{itemize}
\item Standard library:  40\%
\item REPL:              20\%
\item Metaprogramming:   10\%
\item Annotations:       10\%
\item Testing:           10\%
\item Hot swapping:       5\%
\item Typing:             5\%
\end{itemize}
